Generalized spaces for constructive algebra

### 2020-04-08 paper-herrsching2019/main.tex
Reformatted introduction

\begin{abstract}
XXX This contribution gives an introduction to the theory of locales,
\end{abstract}
\body

\tableofcontents
\bigskip

\noindent
The purpose of this contribution is to give a coherent account of a particular
narrative which links locales, geometric theories, sheaf semantics and
constructive commutative algebra. We are hoping to convey a firm grasp of three
ideas:

\begin{enumerate}
\item Locales are a kind of space in which opens instead of points are
fundamental.
% Any propositional geometric theory gives rise to a locale of models.
\item Sheaf semantics allows us to explore mathematical objects from custom-tailored
mathematical universes.
\item Without loss of generality, any reduced ring is a field.
\end{enumerate}

Unlike some other texts on locales or sheaves, this text is set in a
constructive (but impredicative) metatheory. We do not use the law of excluded
middle nor any version of the axiom of choice, and we also do not adopt any
nonclassical principles. We freely use powersets, but more for linguistic
convenience than by necessity.

For readers familiar with the themes of this contribution, we would like to
make a couple of remarks.
\begin{enumerate}
\item We present a self-contained proof of the simple
version of Barr's theorem as Theorem~\ref{thm:baby-barr}. For reducing
intuitionistic provability to geometric provability, we employ (what amounts
to) the syntactic site, similar to the refined completeness result
by Thierry Coquand~\cite{coquand:site} but closer to the standard description
of the syntactic site. For reducing classical provability to intuitionistic
provability, we employ a variant of the double negation translation. We were
not able to track down a reference to this approach in the context of
infinitary logic, but the basic idea has long been known and is summarized, for
instance, in Ref.~\refcite{aczel:russell-prawitz}.
\item In Section~\ref{sect:sheaves}, where we introduce sheaves, we should
probably also introduce sites. They would fit the narrative well and allow for
the construction of generic models of arbitrary geometric theories and not only
propositional ones. They would also allow us to cut down on the metatheory,
sitting comfortably within predicative environments such as arithmetic
universes~\cite{maietti:au,vickers:sketches} or~\textsc{czf}.
We do not, however, mainly to not presume
prerequisites in category theory.
\item The technique presented in Section~\ref{sect:appl} can be
regarded as a first-order (and, if desired, higher-order) coating of an
underlying geometric theory. We propose to take this coating seriously; in
particular to make good use of surprising nongeometric sequents validated by the
generic model; and, for ease of use by algebraists, to cast it in semantic terms,
even though it is all syntax behind the scenes.\footnote{By now it is an established idea in constructive algebra that we should
consider the theory of prime filters instead of the prime filters themselves,
hoping that, for instance, a given mathematical proof of the statement~``no
prime filter contains~$x$'' can also be cast in the
geometric language of the theory, which then yields the nilpotency of~$x$
constructively. If the given proof uses only geometric reasoning, this is
immediate; if it uses intuitionistic reasoning, it can be compiled down (for
instance precisely by the technique we use); and if it uses classical reasoning, a
version of Barr's theorem might be applicable. The elusive actual prime filters
of the ring are replaced with the concrete and purely syntactical generic prime
filter.

This approach misses that the generic model is interesting on its own and in
particular that it satisfies peculiar nongeometric sequents such as ``nonunits
are nilpotent'' and ``any ideal is \notnot finitely generated''. In this sense,
the generic model is better than the actual models it aims to replace, and we
believe that potential applications of this insight should be properly
explored.}
\end{enumerate}

@@

@@

\begin{definition}A \emph{geometric theory} consists of
\begin{enumerate}
  \item a set of sorts: $X$, $Y$, $Z$, \ldots
  \item a set of function symbols: $f : X \times Y \to Z$, \ldots
  \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
  \item a set of geometric sequents as axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
\end{enumerate}
A \emph{geometric sequent} (in some context~$x_1\?X_1,\ldots,x_n\?X_n$) is a
formula built using only the ingredients ${=}\ {\top}\ {\wedge}\ {\bot}\
{\vee}\ {\bigvee}\ {\exists}$ and the relation symbols (but no ${\Rightarrow}\
{\forall}$). We follow the usual convention that negation is understood as
syntactic sugar for implication to~$\bot$. The symbol~``$\bigvee$'' refers to disjunction of arbitrary
set-indexed families of formulas. The theory \emph{proves} a sequent~$\sigma$ if and
only if there exists a \emph{derivation} of~$\sigma$, where the class of
derivations is inductively generated by the axioms and the rules displayed in
Table~\ref{table:geometric-logic}.\footnote{Because of the infinitary
disjunctions, some care is needed to not ``built in'' the axiom of choice in
the definition of derivations. More specifically, we allow inhabited sets of
subdeductions instead of requiring that subdeductions are given by a function.
Details on the implementation in~\textsc{czf} can be found in
Ref.~\cite[Definition~5.2]{rathjen:barr}.}
\end{definition}

@@

Even though superficially similar, geometric theories play a
substantially different role than formal systems such as Peano arithmetic or
Zermelo--Fraenkel set theory. There are a number of notable differences:
\begin{enumerate}
\item Formal systems are typically of foundational interest and can be
fruitfully employed as metatheories. Geometric theories are more interesting
for applications, for instance for presenting spaces. This aspect is discussed
in Section~\ref{sect:presenting-frames}.
\item Formal systems typically come with an intended model; geometric theories
do not. For instance, the geometric theory of groups has all groups as models;
when we are writing down the axioms of a group, we are not setting out to
capture the properties of any one specific group.
\item Formal systems are typically required to be recursively
axiomatizable. Many interesting geometric theories are not, and indeed contain
an uncountable number of sorts, function symbols, relation symbols or axioms. Geometric theories also allow for arbitrary
set-indexed disjunctions; for this to make sense, the study of geometric
theories can only be carried out in a sufficiently rich background theory.
\item For formal systems, typically a version of Gödel's completeness theorem
holds: A formula is derivable iff it holds in all models. In contrast,
geometric theories can be consistent yet do not admit any set-based models. An
example is the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
reviewed below.
\item Perhaps the most important difference to formal systems is that geometric
theories often depend on given mathematical objects. For instance, for each
ring~$A$, there is the geometric theory of prime ideals of~$A$. This theory
will play a crucial role in Section~\ref{sect:appl}.
\end{enumerate}

@@

\paragraph{Extracting constructive content from classical proofs.}\label{par:barr}
Geometric logic strikes a fine balance between being sufficiently expressive in order to
have rich applications and also being sufficiently restrictive so that \emph{Barr's
theorem} holds. This theorem is concerned with eliminating classical reasoning
from proofs of geometric sequents, and it comes in a simple yet still useful
and a more sophisticated version:
\begin{enumerate}
\item If infinitary classical logic proves a geometric sequent, then so does
geometric logic.
\item If a specific version of extensional type theory with the law of excluded
middle and the axiom of choice proves a geometric sequent, then so does
geometric logic.
\end{enumerate}
Barr's theorem has ramifications to constructive algebra, as pointed out by
Gavin Wraith~\cite{wraith:intuitionistic-algebra}.
For instance, the \emph{Nullstellensatz} is the statement that for any
system~$f_1,\ldots,f_n \in K[X_1,\ldots,X_m]$ of polynomials over an
algebraically closed field~$K$, either the polynomials~$f_i$ have a common zero
or there are polynomials~$g_i$ such that~$1 = \sum_i f_i g_i$ (such an
expression is an \emph{algebraic witness} of the nonexistence of a common
zero).

@@

However, knowing that these proofs exist platonically is still worthwhile, as
we can be assured that looking for these proofs is not bound to fail. Indeed,
for the example of the Nullstellensatz, a constructive proof is given
as~\cite[Theorem~9.7]{lombardi-quitte:constructive-algebra}.

@@

As a consequence, a propositional geometric theory consists just of a set of
nullary relation symbols and a set of axioms. Associated to any such
theory~$\TT$ is its \emph{Lindenbaum algebra}: This is the partially ordered
set of the formulas over the signature of~$\TT$ modulo~$\TT$-provable
equivalence, equipped with the ordering given by~$[\varphi] \preceq [\psi]$ iff~$\TT$
proves~$\varphi \vdash \psi$.

@@

\begin{proposition}\label{prop:at-points}
Let~$\TT$ be a geometric theory. Let~$\TT_0$ be the same
theory as~$\TT$ but without any axioms. Let~$M$ be a sheaf model
of~$\TT_0$ on a locale~$X$ (hence just a \emph{sheaf structure}) and consider
the structures~$M_x$ obtained by computing the stalks at points~$x \in \Pt(X)$.
Let~$\sigma$ be a geometric sequent.
Then:
\begin{enumerate}
\item If~$M$ validates~$\sigma$, then so does~$M_x$, for any point~$x \in \Pt(X)$.
\item The converse holds if~$X$ is spatial.
\end{enumerate}
\end{proposition}

@@

{\scriptsize\emph{Hint.} Employ countably many relation symbols~$R_n
\hookrightarrow X \times \cdots \times X$, where~$R_n(x_1,\ldots,x_n)$ should
signify that the elements~$x_1,\ldots,x_n$ exhaust all elements of~$X$. A
solution is given in~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
\end{alphlist}
\end{exercise}

>>>>

\begin{abstract}
The purpose of this contribution is to give a coherent account of a particular
narrative which links locales, geometric theories, sheaf semantics and
constructive commutative algebra. We are hoping to convey a firm grasp of three
ideas:
(1)~Locales are a kind of space in which opens instead of points are
fundamental.
(2)~Sheaf semantics allows us to explore mathematical objects from custom-tailored
mathematical universes.
(3)~Without loss of generality, any reduced ring is a field.
\end{abstract}
\body

\tableofcontents

@@

\paragraph{Remark on the metatheory.}
Unlike some other texts on locales or sheaves, this text is set in a
constructive (but impredicative) metatheory. We do not use the law of excluded
middle nor any version of the axiom of choice, and we also do not adopt any
nonclassical principles. We freely use powersets, but more for linguistic
convenience than by necessity.

\paragraph{Notes for readers familiar with the themes of this contribution.}
We present a self-contained proof of the simple
version of Barr's theorem as Theorem~\ref{thm:baby-barr}. For reducing
intuitionistic provability to geometric provability, we employ (what amounts
to) the syntactic site, similar to the refined completeness result
by Thierry Coquand~\cite{coquand:site} but closer to the standard description
of the syntactic site. For reducing classical provability to intuitionistic
provability, we employ a variant of the double negation translation. We were
not able to track down a reference to this approach in the context of
infinitary logic, but the basic idea has long been known and is summarized, for
instance, in Ref.~\refcite{aczel:russell-prawitz}.

In Section~\ref{sect:sheaves}, where we introduce sheaves, we should
probably also introduce sites. They would fit the narrative well and allow for
the construction of generic models of arbitrary geometric theories and not only
propositional ones. They would also allow us to cut down on the required metatheory,
sitting comfortably within predicative environments such as arithmetic
universes~\cite{maietti:au,vickers:sketches} or~\textsc{czf}.
We do not, however, mainly to not require
prerequisites in category theory.

The technique presented in Section~\ref{sect:appl} can be
regarded as a first-order (and, if desired, higher-order) coating of an
underlying geometric theory. We propose to take this coating seriously; in
particular to make good use of surprising nongeometric sequents validated by the
generic model; and, for ease of use by algebraists, to cast it in semantic terms,
even though it is all syntax behind the scenes.\footnote{By now it is an established idea in constructive algebra that we should
consider the theory of prime filters instead of the prime filters themselves,
hoping that, for instance, a given mathematical proof of the statement~``no
prime filter contains~$x$'' can also be cast in the
geometric language of the theory, which then yields the nilpotency of~$x$
constructively. If the given proof uses only geometric reasoning, this is
immediate; if it uses intuitionistic reasoning, it can be compiled down (for
instance precisely by the technique we use); and if it uses classical reasoning, a
version of Barr's theorem might be applicable. The elusive actual prime filters
of the ring are replaced with the concrete and purely syntactical generic prime
filter.

This approach misses that the generic model is interesting on its own and in
particular that it satisfies peculiar nongeometric sequents such as ``nonunits
are nilpotent'' and ``any ideal is \notnot finitely generated''. In this sense,
the generic model is better than the actual models it aims to replace, and we
believe that potential applications of this insight should be properly
explored.}

@@

\begin{definition}\label{defn:geometric-theory}
A \emph{geometric theory} consists of
\begin{enumerate}
  \item a set of sorts: $X$, $Y$, $Z$, \ldots
  \item a set of function symbols: $f : X \times Y \to Z$, \ldots
  \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
  \item a set of geometric sequents as axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
\end{enumerate}
A \emph{geometric sequent} (in some context~$x_1\?X_1,\ldots,x_n\?X_n$) is a
formula built using only the ingredients ${=}\ {\top}\ {\wedge}\ {\bot}\
{\vee}\ {\bigvee}\ {\exists}$ and the relation symbols (but no ${\Rightarrow}\
{\forall}$). We follow the usual convention that negation is understood as
syntactic sugar for implication to~$\bot$. The symbol~``$\bigvee$'' refers to disjunction of arbitrary
set-indexed families of formulas. The theory \emph{proves} a sequent~$\sigma$ if and
only if there exists a \emph{derivation} of~$\sigma$, where the class of
derivations is inductively generated by the axioms and the rules displayed in
Table~\ref{table:geometric-logic}.\footnote{Because of the infinitary
disjunctions, some care is needed to not ``built in'' the axiom of choice in
the definition of derivations. More specifically, we allow inhabited sets of
subdeductions instead of requiring that subdeductions are given by a function.
Details on the implementation in~\textsc{czf} can be found in
Ref.~\cite[Definition~5.2]{rathjen:barr}.}
\end{definition}

We refer to Ref.~\cite[Section~D1.1]{johnstone:elephant} for a more detailed
version of Definition~\ref{defn:geometric-theory}.

@@

Even though superficially similar, geometric theories play a
substantially different role than formal systems such as Peano arithmetic or
Zermelo--Fraenkel set theory. There are a number of notable differences:
\begin{enumerate}
\item Formal systems are typically of foundational interest and can be
fruitfully employed as metatheories. Geometric theories are more interesting
for applications, for instance for presenting spaces. This aspect is discussed
in Section~\ref{sect:presenting-frames}.
\item Formal systems typically come with an intended model; geometric theories
do not. For instance, the geometric theory of groups has all groups as models;
when we are writing down the axioms of a group, we are not setting out to
capture the properties of any one specific group.
\item Formal systems are typically required to be recursively
axiomatizable. Many interesting geometric theories are not, and indeed contain
an uncountable number of sorts, function symbols, relation symbols or axioms. Geometric theories also allow for arbitrary
set-indexed disjunctions; for this to make sense, the study of geometric
theories can only be carried out in a sufficiently rich background theory.
\item For formal systems, typically a version of Gödel's completeness theorem
holds: A formula is derivable iff it holds in all models. In contrast,
geometric theories can be consistent yet do not admit any set-based models. An
example is the geometric theory of surjections~$\NN \twoheadrightarrow \RR$
reviewed below.
\item Perhaps the most important difference to formal systems is that geometric
theories often depend on given mathematical objects. For instance, for each
ring~$A$, there is the geometric theory of prime filters of~$A$. This theory
will play a crucial role in Section~\ref{sect:appl}.
\end{enumerate}

@@

\paragraph{Extracting constructive content from classical proofs.}\label{par:barr}
Geometric logic strikes a fine balance between being sufficiently expressive in order to
have rich applications and also being sufficiently restrictive so that \emph{Barr's
theorem} holds. This theorem is concerned with eliminating classical reasoning
from proofs of geometric sequents, and it comes in a simple yet still useful
and a more sophisticated version:
\begin{enumerate}
\item If infinitary classical logic proves a geometric sequent, then so does
geometric logic.
\item If a specific version of extensional type theory with the law of excluded
middle and the axiom of choice proves a geometric sequent, then so does
geometric logic.
\end{enumerate}
Barr's theorem has ramifications for constructive algebra, as pointed out by
Gavin Wraith~\cite{wraith:intuitionistic-algebra}.
For instance, the \emph{Nullstellensatz} is the statement that for any
system~$f_1,\ldots,f_n \in K[X_1,\ldots,X_m]$ of polynomials over an
algebraically closed field~$K$, either the polynomials~$f_i$ have a common zero
or there are polynomials~$g_i$ such that~$1 = \sum_i f_i g_i$ (such an
expression is an \emph{algebraic witness} of the nonexistence of a common
zero).

@@

However, knowing that these proofs exist platonically is still worthwhile, as
we can be assured that looking for these proofs is not bound to fail. Indeed,
for the example of the Nullstellensatz, a constructive proof is given
in Ref.~\cite[Theorem~9.7]{lombardi-quitte:constructive-algebra}.

@@

As a consequence, a propositional geometric theory consists just of a set of
nullary relation symbols and a set of axioms. Associated to any such
theory~$\TT$ is its \emph{Lindenbaum algebra}: This is the partially ordered
set of the geometric formulas over the signature of~$\TT$ modulo~$\TT$-provable
equivalence, equipped with the ordering given by~$[\varphi] \preceq [\psi]$ iff~$\TT$
proves~$\varphi \vdash \psi$.

@@

\begin{proposition}\label{prop:at-points}
Let~$\TT$ be a geometric theory without any axioms. Let~$M$ be a sheaf model
of~$\TT$ on a locale~$X$ (hence just a \emph{sheaf structure}) and consider
the structures~$M_x$ obtained by computing the stalks at points~$x \in \Pt(X)$.
Let~$\sigma$ be a geometric sequent.
Then:
\begin{enumerate}
\item If~$M$ validates~$\sigma$, then so does~$M_x$, for any point~$x \in \Pt(X)$.
\item The converse holds if~$X$ is spatial.
\end{enumerate}
\end{proposition}

@@

{\scriptsize\emph{Hint.} Employ countably many relation symbols~$R_n
\hookrightarrow X \times \cdots \times X$, where~$R_n(x_1,\ldots,x_n)$ should
signify that the elements~$x_1,\ldots,x_n$ exhaust all elements of~$X$. A
solution is given in Ref.~\cite[Example~D1.1.7(k)]{johnstone:elephant}.\par}
\end{alphlist}
\end{exercise}

\begin{exercise}{Issues of size}%
Working in~\textsc{izf}, show that the Lindenbaum algebra of a propositional
geometric theory~$\TT$ can be realized as a set, even though the class of
geometric formulas is typically a proper class.\smallskip

\noindent{\scriptsize\emph{Hint.} Use that any geometric formula is provably equivalent
to one of the form~$\bigvee_i \varphi_i$, where each formula~$\varphi_i$ is a
conjunction of nullary relation symbols.\par}
\end{exercise}
